### 2020-06-18 report/sections/background/related.tex
Report: feedback chapter 8
Report: feedback on Ch2

\[
G_\text{Adder} = \trec{
\gcommone{C}{S}{
	\begin{cases}
		\tmsg{Add(int, int)} & :
			\gcommone{S}{C}{\tmsg{Res(int)}}. \trecvar \\
		\tmsg{Quit()} & : 
			\gcommone{S}{C}{\tmsg{Terminate()}}. \tend \\		
	\end{cases}
}} 
\]

@@

\item \textbf{Payload Types:}
The set of message payload data types that the endpoint
sends or expects to receive.
For \cref{fig:bgadderfsm}, this is singleton set
$\{~\texttt{number}~\}$.

@@

\item \textbf{Channel linearity:}
Each transition in the EFSM represents a channel resource.
When the application transitions from some state $S$
to some successor state $S'$, it must no longer
be able to access a reference (e.g. have an alias) to $S$.
\end{itemize}

@@

\subsubsection{Hybrid Session Verification in Java}
Hu and Yoshida implemented a workflow in \cite{Hybrid2016}
that performs \textit{hybrid} verification of 
communication protocol conformance for Java applications.
The ``hybrid'' concept is composed of the two main components below:

@@

\subparagraph{Runtime channel linearity checks}
By exposing send and receive actions as instance
methods, the developer can keep aliases of EFSM state instances
and invoke the IO action more than once. Java does not
have a linear type system, and hence cannot statically
verify linear usage; this means linearity checks need to be
done at runtime.
Each state keeps track of its usage in a private boolean flag a
nd throws an exception when the instance method is called twice.
Similarly, the \texttt{SessionEndpoint} class keeps track of 
whether the connection is open or close,
and throws an exception when program execution exits the 
scope of the session endpoint and a terminal state 
has yet to be reached.\\

@@

\subsubsection{Runtime Monitors in Python}
Neykova targeted the MPST methodology 
for Python programs in \cite{Python2017} 
and proposed to generate \textit{runtime monitors} 
from the EFSM.
These monitors expose APIs for sending and receiving messages,
which is used by the developer in their implementation. 
The runtime monitor is an abstraction between the 
developer's implementation and the actual communication channel, 
and ``executes'' the EFSM internally to ensure protocol conformance. 
When the developer sends a message (with some label and payload) 
using the API, the runtime monitor checks 
whether this send action conforms to the current EFSM state, 
and if so, 
performs the send and advances to the successor state. 
Likewise, when the developer invokes a receive, 
the runtime monitor verifies that this is permitted 
at the current EFSM state before returning the received payload.

@@

\subsubsection{API Generation in PureScript}
King et al. presented an approach in \cite{PureScript2019}
for integrating \textit{multiparty} session types 
into web development using 
the PureScript language \cite{PureScript},
PureScript is a strongly typed functional language
that compiles to JavaScript.
This proposal takes advantage of its expressive type system to 
provide \textit{static} guarantees for protocol conformance.
We outline the main components of their EFSM encoding:

@@

\subparagraph{Transitions as instances of type classes} 
By encoding states as data types, 
valid EFSM transitions are encoded as 
\textit{instances} of the type classes. 
If \texttt{S1} is an output state, 
sending an \texttt{Int} to \texttt{Svr} with 
successor state \texttt{S2}, the PureScript encoding would be
\lstonelineps{instance SendS1 :: Send Svr S1 S2 Int}.
Because of the functional dependency, 
the developer cannot instantiate an invalid transition 
(e.g. \lstonelineps{Send Svr S1 S3 Bool}, 
since \texttt{S1} uniquely determines the other type parameters. 
PureScript statically resolves these type constraints at
compile-time.

@@

This proposal is relevant to the problem we are tackling.
In particular, the final point about guaranteeing channel linearity
by construction is something we can build upon in our solution,
since linear types is a feature left to be desired in
``mainstream'' programming languages.
We also observe the challenges of session-typed GUI
programming, as shown by the careful choice 
made in \cite{PureScript2019} 
to use the \textit{Concur UI} framework \cite{ConcurUI} which 
builds UI elements sequentially to model sequential sessions.
Not doing so would make channel linearity violation possible 
(e.g. by binding a send transition to a button click event,
the button may still be active after the transition is complete;
clicking on it again will reuse the channel resource).
Whilst \cite{PureScript2019} supports multiparty session types,
the communication protocol must describe all interactions
to go through a centralised server role to respect the
network topology enforced by WebSocket transport.

@@

\subsubsection{Session-Typed GUI Programming using Links}
Fowler presented the first formal integration
of session typing and GUI programming in \cite{MVU2020}.
His work formalised the \textit{Model-View-Update} (MVU)
architectural pattern for GUI development.
An MVU application features a \textit{model} encapsulating application
state, a \textit{view function} rendering the state on the 
user interface, and an
\textit{update function} handling
\textit{messages} produced by the
rendered model to produce a new model.
Fowler introduces the concept of model types
to express type dependencies between these components:
a \emph{model type} uniquely defines a \textit{view} function,
set of \textit{messages} and \textit{update function} --
rather than producing a new model, the update function 
defines valid transitions to other model types.
This approach guarantees channel linearity for GUI programming:
the update function transitions to a successor model, which only
renders an interface with the set of messages unique to that model,
so channel actions from the predecessor model are not rendered,
which makes reuse impossible.
The core MVU calculus is also extended with \textit{commands}
and \textit{linearity} required to express communication
actions in session type theory.

\cite{MVU2020} is implemented using
the \textit{Links} web programming language
\cite{LINKS}. Links is a strict, typed, functional language
that can be used to write full-stack web applications;
and similar to PureScript, it is compiled to JavaScript.
As it has a linear type system, session types can be
implemented using native language features, since the
language can monitor linear channel usage.
Links also supports error handling for session types,
as introduced in \cite{Exceptional}; 
given side-effects must be made explicit
in functional programming, \cite{Exceptional} introduces
a \textit{session cancellation} operator to ensure
that channel resources are closed after handling
side-effects, such as errors.

>>>>

\[
G_\text{Adder} = \trec{
\gcommone{C}{S}{
	\begin{cases}
		\tmsg{Add(int, int)} & :
			\gcommone{S}{C}{\tmsg{Res(int)}}. ~\trecvar \\
		\tmsg{Quit()} & : 
			\gcommone{S}{C}{\tmsg{Terminate()}}.~ \tend \\		
	\end{cases}
}} 
\]

@@

\item \textbf{Payload Types:}
The set of message payload data types that the endpoint
sends or expects to receive.
For \cref{fig:bgadderfsm}, this is singleton set
$\{~\texttt{int}~\}$.

@@

\item \textbf{Channel linearity:}
Each transition in the EFSM represents one channel resource.
The application transitions from some state $S$
to some successor state $S'$
once \textit{and only once},
after which it must no longer
be able to access a reference (e.g. have an alias) to $S$.
\end{itemize}

@@

\subsubsection{Hybrid Session Verification in Java \cite{Hybrid2016}}
Hu and Yoshida implemented a workflow
that performs \textit{hybrid} verification of 
communication protocol conformance for Java applications.
The ``hybrid'' concept is composed of the two main components below:

@@

\subparagraph{Runtime channel linearity checks}
By exposing send and receive actions as instance
methods, the developer can keep aliases of EFSM state instances
and invoke the IO action more than once. Java does not
have a linear type system, and hence cannot statically
verify linear usage; this means linearity checks need to be
done at runtime.
Each state keeps track of its usage in a private boolean flag 
and throws an exception when the instance method is called twice.
Similarly, the \texttt{SessionEndpoint} class keeps track of 
whether the connection is open or close,
and throws an exception when program execution exits the 
scope of the session endpoint and a terminal state 
has yet to be reached.\\

@@

\subsubsection{Runtime Monitors in Python \cite{Python2017} }
Neykova targeted the MPST methodology 
for Python programs
and proposed to generate \textit{runtime monitors} 
from the EFSM.
These monitors expose APIs for sending and receiving messages,
which is used by the developer in their implementation. 
The runtime monitor is an abstraction between the 
developer's implementation and the actual communication channel, 
and ``executes'' the EFSM internally to ensure protocol conformance. 
When the developer sends a message (with some label and payload) 
using the API, the runtime monitor checks 
whether this send action conforms to the current EFSM state, 
and if so, 
performs the send and advances to the successor state. 
Likewise, when the developer invokes a receive, 
the runtime monitor verifies that this is permitted 
at the current EFSM state before returning the received payload.

@@

\subsubsection{API Generation in PureScript \cite{PureScript2019}}
King et al. presented an approach
for integrating \textit{multiparty} session types 
into web development using 
the PureScript language \cite{PureScript},
PureScript is a strongly typed functional language
that compiles to JavaScript.
This proposal takes advantage of its expressive type system to 
provide \textit{static} guarantees for protocol conformance.
We outline the main components of their EFSM encoding:

@@

\subparagraph{Transitions as instances of type classes} 
By encoding states as data types, 
valid EFSM transitions are encoded as 
\textit{instances} of the type classes. 
If \texttt{S1} is an output state, 
sending an \texttt{Int} to \texttt{Svr} with 
successor state \texttt{S2}, the PureScript encoding would be
\lstonelineps{instance SendS1 :: Send Svr S1 S2 Int}.
Because of the functional dependency, 
the developer cannot instantiate an invalid transition 
(e.g. \lstonelineps{Send Svr S1 S3 Bool}, 
since \texttt{S1} uniquely determines the other type parameters. 
PureScript statically resolves these type constraints at
compile time.

@@

This proposal is relevant to the problem we are tackling.
In particular, the final point about guaranteeing channel linearity
by construction is something we can build upon in our solution,
since linear types is a feature left to be desired in
``mainstream'' programming languages.
We also observe the challenges of session-typed GUI
programming, as shown by the careful choice 
made in \cite{PureScript2019} 
to use the \textit{Concur UI} framework \cite{ConcurUI} which 
builds UI elements sequentially to model sequential sessions.
Not doing so would make channel linearity violation possible 
(e.g. by binding a send transition to a button click event,
the button may still be active after the transition is complete;
clicking on it again will reuse the channel resource).
Whilst King et al. \cite{PureScript2019} supports 
multiparty session types,
the communication protocol must describe all interactions
to go through a centralised server role to respect the
network topology enforced by WebSocket transport.

@@

\subsubsection{Session-Typed GUI Programming 
using Links \cite{MVU2020}}
Fowler presented the first formal integration
of session typing and GUI programming .
His work formalised the \textit{Model-View-Update} (MVU)
architectural pattern for GUI development.
An MVU application features a \textit{model} encapsulating application
state, a \textit{view function} rendering the state on the 
user interface, and an
\textit{update function} handling
\textit{messages} produced by the
rendered model to produce a new model.
Fowler introduces the concept of model types
to express type dependencies between these components:
a \emph{model type} uniquely defines a \textit{view} function,
set of \textit{messages} and \textit{update function} --
rather than producing a new model, the update function 
defines valid transitions to other model types.
This approach guarantees channel linearity for GUI programming:
the update function transitions to a successor model, which only
renders an interface with the set of messages unique to that model,
so channel actions from the predecessor model are not rendered,
which makes reuse impossible.
The core MVU calculus is also extended with \textit{commands}
and \textit{linearity} required to express communication
actions in session type theory.

Fowler's proposal \cite{MVU2020} is implemented using
the \textit{Links} web programming language
\cite{LINKS}. Links is a strict, typed, functional language
that can be used to write full-stack web applications;
and similar to PureScript, it is compiled to JavaScript.
As it has a linear type system, session types can be
implemented using native language features, since the
language can monitor linear channel usage.
Links also supports error handling for session types,
as introduced in \cite{Exceptional}; 
given side-effects must be made explicit
in functional programming,
Fowler et al. \cite{Exceptional} introduces
a \textit{session cancellation} operator to ensure
that channel resources are closed after handling
side-effects, such as errors.

### 2020-06-18 report/sections/background/sessiontypes.tex
Report: complete background
Report: review chapter 4
Report: feedback on Ch2

\section{Session Types}

@@

The process algebra $\pi$-calculus 
introduced by Milner in \cite{Milner1999} 
formalises the message passing abstraction in terms of 
the basic building blocks of sending and receiving processes,
along with inductively defined continuation processes. 
The composition of these primitives allow us to 
describe more complex communication sessions.
\textit{Session types} define the typing discipline 
for the $\pi$-calculus and provide reliability guarantees 
for communication sessions; 
the latter addresses a key challenge when 
reasoning about the correctness of distributed systems.

Practical application of session types in software
engineering range from
developing languages providing native session type support 
\cite{ATS2016} to implementing session types 
in existing programming languages across different paradigms.
Implementations of the latter approach differ by 
how they leverage the design philosophy and features 
provided by the programming language.
For example, King et al. leveraged the 
expressive type system of PureScript to perform 
\textit{static} session type checking in \cite{PureScript2019},
whilst Neykova and Yoshida introduced dynamic approaches to 
check the conformance of Python programs 
with respect to session types in \cite{Python2017}.
We discuss these related work, among others, in \cref{section:bgrelated}.

@@

\begin{itemize}
\item $\mathbf{0}$ is the nil process and represents inactivity.
\item $\pout{u}{v}$ is the output process that will send value $v$ on $u$.
\item $\pin{u}{x}.P$ is the input process that,
 upon receiving a message on $u$, 
 will bind the message to $x$ and 
 carry on executing $P$ under this binding.
\item $P\mid Q$ represents the parallel composition of 
processes executing simultaneously.
\item $!P$ represents the parallel composition of 
\textit{infinite} instances of $P$; 
more specifically, $!P \equiv P \mid {!P}$.
\item $(\nu a)~P$ represents a name restriction 
where any occurrence of $a$ in $P$ is local
 and will not interfere with other names outside the scope of $P$.
\end{itemize}

@@

The operational semantics model the interaction 
between parallel processes. 
Whilst \cite{C406Lecture} presents the full operational semantics, 
we highlight the \rulename{Comm} reduction rule which 
specifically models message passing:
if the parallel composition of an input process and output process 
share the same name, the composition reduces to the 
continuation of the input process, 
substituting the variable $x$ with the message received $v$. 
We omit the definitions of substitution, free variables and free names, 
$\alpha$-equivalence and structural congruence; 
the interested reader may refer to \cite{C406Lecture}.

@@

\begin{prooftree}
\AxiomC{$
\forall i \in I \left(
\vec x_i = x_1, \dots, x_n
\qquad
\vec U_i = U_1, \dots, U_n
\qquad
\Gamma, x_1: U_1, \dots, x_n: U_n \vdash P_i : T_i
\right)$}
\RightLabel{\rulename{Ty-Bra}}
\UnaryInfC{$
\Gamma \vdash \bra{p}{l_i(\vec x_i): P_i}{i\in I} : 
\tbra{p}{l_i(\vec U_i):S_i}{i\in I}
$}
\end{prooftree}

@@

\[
\begin{array}{rl}
\mathcal{M} ::= & \ptprocess{p_1}{P_1} ~\mid~
\ptprocess{p_2}{P_1} ~\mid~\dots~\mid
\ptprocess{p_n}{P_n}
\end{array}
\]

the same cannot be said for the binary session typing discipline,
particularly with respect to duality.
The same notion of duality does not extend to the 
decomposition of multiparty interactions into 
multiple binary sessions: 
\cite{C406Lecture} and \cite{MPST} both present 
counterexamples of well-typed binary sessions that, 
when composed to represent a multiparty session, 
results in communication errors thus 
violating guarantees of well-typed sessions.

Honda et al. presents {multiparty session types} 
in \cite{MPAST} to extend the 
binary session typing discipline for 
sessions involving more than 2 participants, 
whilst redefining the notion of compatibility 
in this multiparty context. 
\textit{Multiparty session types} (MPST)
are defined in terms a \textit{global type},
which provides a bird's eye view of the communication protocol
describing the interactions between pairs of participants. 
\cref{fig:mpst} defines the syntax of global types
inspired by \cite{MPST};
we omit message payload types for brevity.

@@

G ::= & & \text{Global Types} \\
     & \tend & \text{Termination} \\
\mid & \gcomm{p}{q}{l_i: G_i}{i \in I} & \text{Message Exchange} \\
\mid & \trec{G} & \text{Recursive Type} \\
\mid & \trecvar & \text{Type Variable} \\
\end{array}
\]
\singlespacing
\caption{Syntax of Global Types}
\label{fig:mpst}
\end{figure}

@@

More formally, projection can be interpreted as a 
\textit{partial function} 
${\upharpoonright} :: G \times \mrole{p} \rightharpoonup S$,
as the projection for a participant may be undefined for 
an ill-formed global type; 
\cite{C406Lecture} presents examples of where this is the case, 
and \cite{MPST} presents the formal definition of projection.
We focus on the definition of projection for the 
message exchange global type given below:

@@

The third case describes that,
for a participant not involved in the communication,
the projections of the continuations $\proj{G_i}{r}$
do not necessarily have to be the same, but they
need to be ``compatible''. 
The $\mergeop$ syntax denotes the \textit{merging operator}.
It is a partial function on local types allows us to construct
a projection that is compatible with the projections on
the continuations.
We use the \textit{inductive} definition presented by
Scalas and Yoshida in \cite{LessIsMore}, shown in \cref{fig:merge}.

>>>>

\section{Session Types}
\label{section:sessiontypes}

@@

The process algebra $\pi$-calculus 
introduced by Milner in \cite{Milner1999} 
formalises the message passing abstraction in terms of 
the basic building blocks of sending and receiving processes.
The composition of these primitives allow us to 
describe more complex communication sessions.
\textit{Session types} define the typing discipline 
for the $\pi$-calculus and provide reliability guarantees 
for communication sessions; 
the latter addresses a key challenge when 
reasoning about the correctness of distributed systems.

Practical application of session types in software
engineering range from
developing languages providing native session type support 
\cite{ATS2016} to implementing session types 
in existing programming languages across different paradigms.
Implementations of the latter approach differ by 
how they leverage the design philosophy and features 
provided by the programming language.
For example, King et al. \cite{PureScript2019} leveraged the 
expressive type system of PureScript to perform 
\textit{static} session type checking,
whilst Neykova and Yoshida \cite{Python2017} 
introduced dynamic approaches to 
check the conformance of Python programs 
with respect to session types.
We discuss these related work, among others, in \cref{section:bgrelated}.

@@

\begin{itemize}
\item $\mathbf{0}$ is the nil process and represents inactivity.
\item $\pout{u}{v}$ is the output process that will send value $v$ on $u$.
\item $\pin{u}{x}.P$ is the input process that,
 upon receiving a message on $u$, 
 will bind the message to $x$ and 
 carry on executing $P$ under this binding.
\item $P\mid Q$ represents the parallel composition of 
processes executing simultaneously.
\item $!P$ represents the parallel composition of 
\textit{infinite} instances of $P$; 
more specifically, $!P \equiv P \mid {!P}$.

$\equiv$ represents the \textit{structural congruence} 
equivalence relation
on processes, meaning that $!P$ is indistinguishable from 
$P \mid {!P}$. We inherit the definition of structural congruence from
\cite{C406Lecture}.

\item $(\nu a)~P$ represents a name restriction 
where any occurrence of $a$ in $P$ is local
 and will not interfere with other names outside the scope of $P$.
\end{itemize}

@@

The operational semantics model the interaction 
between parallel processes.
We inherit the full operational semantics presented by
Yoshida \cite{C406Lecture}, and
highlight the \rulename{Comm} reduction rule which 
specifically models message passing:
if the parallel composition of an input process and output process 
share the same name, the composition reduces to the 
continuation of the input process, 
substituting the variable $x$ with the message received $v$. 
We omit the definitions of substitution, free variables and free names, 
$\alpha$-equivalence and structural congruence; 
the interested reader may refer to \cite{C406Lecture}.

@@

\begin{prooftree}
\AxiomC{$
\forall i \in I.\left(
\vec x_i = x_1, \dots, x_n
\qquad
\vec U_i = U_1, \dots, U_n
\qquad
\Gamma, x_1: U_1, \dots, x_n: U_n \vdash P_i : T_i
\right)$}
\RightLabel{\rulename{Ty-Bra}}
\UnaryInfC{$
\Gamma \vdash \bra{p}{l_i(\vec x_i): P_i}{i\in I} : 
\tbra{p}{l_i(\vec U_i):S_i}{i\in I}
$}
\end{prooftree}

@@

\[
\begin{array}{rl}
\mathcal{M} ::= & \ptprocess{p_1}{P_1} ~\mid~
\ptprocess{p_2}{P_1} ~\mid~\dots~\mid~
\ptprocess{p_n}{P_n}
\end{array}
\]

the same cannot be said for the binary session typing discipline,
particularly with respect to duality.
The same notion of duality does not extend to the 
decomposition of multiparty interactions into 
multiple binary sessions: 
Yoshida and Lorenzo \cite{MPST} presents
counterexamples of well-typed binary sessions that, 
when composed to represent a multiparty session, 
results in communication errors thus 
violating guarantees of well-typed sessions.

Honda et al. \cite{MPAST} presents {multiparty session types} 
to extend the 
binary session typing discipline for 
sessions involving more than 2 participants, 
whilst redefining the notion of compatibility 
in this multiparty context. 
\textit{Multiparty session types} (MPST)
are defined in terms a \textit{global type},
which provides a bird's eye view of the communication protocol
describing the interactions between pairs of participants. 
\cref{fig:globaltypes} defines the syntax of global types
inspired by \cite{MPST};
we omit message payload types for brevity.

@@

G ::= & & \text{Global Types} \\
     & \tend & \text{Termination} \\
\mid & \gcomm{p}{q}{l_i: G_i}{i \in I} & \text{Message Exchange} \\
\mid & \trec{G} & \text{Recursive Type} \\
\mid & \trecvar & \text{Type Variable} \\
\end{array}
\]
\singlespacing
\caption{Syntax of Global Types}
\label{fig:globaltypes}
\end{figure}

@@

More formally, projection can be interpreted as a 
\textit{partial function} 
${\upharpoonright} :: G \times \mrole{p} \rightharpoonup S$,
as the projection for a participant may be undefined for 
an ill-formed global type; 
Yoshida \cite{C406Lecture} presents examples of where this is the case, 
and Yoshida and Lorenzo \cite{MPST} 
presents the formal definition of projection.
We focus on the definition of projection for the 
message exchange global type given below:

@@

The third case describes that,
for a participant not involved in the communication,
the projections of the continuations $\proj{G_i}{r}$
do not necessarily have to be the same, but they
need to be ``compatible''. 
The $\mergeop$ syntax denotes the \textit{merging operator}.
It is a partial function on local types allows us to construct
a projection that is compatible with the projections on
the continuations.
We use the \textit{inductive} definition presented by
Scalas and Yoshida in \cite{LessIsMore} Definition 3.3,
shown in \cref{fig:merge}.
